---
title: "Google Data Analytics Capstone"
author: "Silas Njoroge"
date: "2025-03-24"
output:
  html_document: default
---

The Data.
The project is an analysis on the Students Performance data in R studio. Data obtained from Kaggle.
The dataset includes the following variables:
1.	Hours Studied – Number of hours a student studied.
2.	Previous Scores – Previous academic performance (e.g., test scores).
3.	Extracurricular Activities – Whether the student participates in extracurricular activities (Yes/No).
4.	Sleep Hours – Number of hours of sleep per night.
5.	Sample Question Papers Practiced – Number of sample question papers the student practiced.
6.	Performance Index – The outcome or dependent variable measuring student performance.


The Business Questions

1.	Performance Analysis:
  o	What is the correlation between the number of hours studied and the performance index?
  o	How much do previous scores predict current performance?
  o	Do students who practice more sample question papers tend to perform better?
2.	Behavior and Lifestyle:
  o	How does sleep quality (hours of sleep) affect student performance?
  o	Do students who engage in extracurricular activities perform better or worse than those who don't?
  o	Is there an optimal balance between sleep, study time, and extracurricular participation?
3.	Interventions and Recommendations:
  o	What combination of study hours, sleep, and extracurricular activities maximizes student performance?
  o	Are students with poor previous scores improving with more study time or other interventions?
4.	Segmentation and Grouping:
  o	Can students be grouped based on similar study habits and performance outcomes?
  o	Are there patterns among high-performing versus low-performing students?
5.	Predictive Modeling:
  o	Can student performance be predicted accurately based on these factors?
  o	What is the most influential factor in predicting student success?


Process
  Data cleaning and wrangling
  Exploratory data analysis (EDA)
  Machine learning (regression, classification)
  Data visualization (trends, correlations)
  Business insights and recommendations


```{r}
#Data cleaning and wrangling
getwd()
my_df <- read.csv("Student_Performance.csv")

##Exploratory Data Analysis (EDA)
head(my_df)

```
```{r}
hist(my_df$Hours.Studied, main = "Histogram of Hours.Studied")

```
#Analysis of Histogram of Hours Studied
The data is unifromly distributed, no skewness.
Apart from 1-2 hours, all the other bars of the histogram do not touch. 
This is because each bar represents a distinct, separate value.
For that reason, further inspect of each bar will be necessary to view the distribution within each hour using Box Plots
```{r}
# Load the necessary library
library(ggplot2)

# Create the box plot
ggplot(my_df, aes(x = factor(Hours.Studied), y = Performance.Index)) +
  geom_boxplot() +
  labs(x = "Hours Studied", y = "Performance Index", title = "Performance Index by Hours Studied") +
  theme_minimal()

```
#Trend Analysis of Box Plots:
1. Positive Correlation: There appears to be a general positive correlation between the number of hours studied and the performance index. Higher hours studied tend to result in higher median performance scores.
2. Slight Increase in Variation: The variation (spread) in performance increases slightly as hours studied increase. This suggests that while studying more generally improves performance, some students who study more still perform poorly — indicating other influencing factors like study quality or test-taking skills.
3. Outliers: There are a few noticeable outliers, especially at higher hours studied, where some students perform significantly better or worse than the typical range.
4. Plateau Effect: After a certain point (around 7–8 hours), the median performance index seems to stabilize, indicating diminishing returns for additional hours studied.


```{r}
hist(my_df$Sleep.Hours, main = "Histogram of Sleep.Hours")

```
```{r}
# Create the box plot
ggplot(my_df, aes(x = factor(Sleep.Hours), y = Performance.Index)) +
  geom_boxplot() +
  labs(x = "Sleep Hours", y = "Performance Index", title = "Performance Index by Sleep Hours")
```
#Trend Analysis of Sleep Hours
1.Moderate Positive Correlation: There seems to be a moderate relationship between sleep hours and performance. Students who sleep around 7–9 hours tend to have higher and more consistent performance scores.
2.Decreased Performance with Less Sleep: Performance is more inconsistent and generally lower for students who sleep fewer than 6 hours. This suggests that insufficient sleep may negatively impact performance.
3. Diminishing Returns: Beyond 9 hours, the performance index seems to stabilize or even slightly decrease, indicating that excessive sleep may not lead to improved performance.
4.Higher Variability with Lower Sleep: Students with very low sleep hours (around 4–5 hours) show greater variability in performance, suggesting that some students may still perform well despite lack of sleep — possibly due to other compensating factors (like prior knowledge or better study habits).

```{r}
hist(my_df$Sample.Question.Papers.Practiced, main = "Histogram of Sample.Question.Papers.Practiced")

```
```{r}
# Create the box plot
ggplot(my_df, aes(x = factor(Sample.Question.Papers.Practiced), y = Performance.Index)) +
  geom_boxplot() +
  labs(x = "Sample.Question.Papers.Practiced", y = "Performance Index", title = "Performance Index by Sample Question Papers Practiced")
```
#Trend Analysis: 
Performance vs. Sample Questions Practiced
Upward Trend: There's a visible upward trend in the median performance index as the number of sample question papers practiced increases. This indicates that practicing more questions tends to result in better performance.

Reduced Variability: Students who practiced more sample questions (e.g., 4 or 5) not only performed better on average but also showed less variability, suggesting more consistent high performance.

Low Practice, Lower Scores: Students who practiced fewer or no sample questions often had lower performance indices, and the spread of scores was wider—indicating greater inconsistency and risk of underperformance.

Few Outliers: There are some lower-performing outliers among students who practiced more, which could reflect individual differences like test anxiety or gaps in understanding.








```{r}

plot(my_df$Hours.Studied, my_df$Performance.Index,
     xlab = "Hours Studied", ylab = "Performance Index",
     main = "Scatter Plot: Hours Studied vs. Performance Index")

# You can add regression line
abline(lm(my_df$Performance.Index ~ my_df$Hours.Studied), col = "red")

```
1.	Performance Analysis:

#What is the correlation between the number of hours studied and the performance index?
```{r}
library(ggplot2)

ggplot(my_df, aes(x = Hours.Studied, y = Performance.Index)) +
  geom_point() +
  geom_smooth(method = "lm", col = "red") +
  labs(title = "Previous Scores vs. Performance Index",
       x = "Previous Scores",
       y = "Performance Index")


```
```{r}
# Calculate the correlation coefficient (r)
r <- cor(my_df$Hours.Studied, my_df$Performance.Index)
cat("Correlation coefficient (r):", r, "\n")

# Calculate R-squared (r^2)
r_squared <- r^2
cat("Coefficient of determination (R²):", r_squared, "\n")

```

Correlation coefficient (r): 0.374 (rounded to 3 decimal places)  
Coefficient of determination (R²): 0.140

# Interpretation:
r = 0.374: There is a moderate positive relationship between hours studied and performance index.
R² = 0.140: About 14%  of the variation in the performance index can be explained by the number of hours studied.

#How much do previous scores predict current performance?

```{r}
# Calculate the correlation coefficient (r)
r <- cor(my_df$Previous.Scores, my_df$Performance.Index)
cat("Correlation coefficient (r):", r, "\n")

# Calculate R-squared (r^2)
r_squared <- r^2
cat("Coefficient of determination (R²):", r_squared, "\n")

```

#Interpretation:
Very Strong Positive Correlation: A correlation of 0.915 indicates a very strong linear relationship between Previous Scores and Current Performance Index.

R² = 0.838: Approximately 84% of the variation in performance can be explained by previous scores alone.

#Do students who practice more sample question papers tend to perform better?
```{r}
# Calculate the correlation coefficient (r)
r <- cor(my_df$Sample.Question.Papers.Practiced, my_df$Performance.Index)
cat("Correlation coefficient (r):", r, "\n")

# Calculate R-squared (r^2)
r_squared <- r^2
cat("Coefficient of determination (R²):", r_squared, "\n")

```
#Interpretation:
Very Weak Positive Correlation: The correlation between the number of sample question papers practiced and performance is very low.
R² = 0.00187: Less than 2% of the variation in performance index is explained by the number of papers practiced.

2.	Behavior and Lifestyle:

#How does sleep quality (hours of sleep) affect student performance?

#Do students who practice more sample question papers tend to perform better?
```{r}
# Calculate the correlation coefficient (r)
r <- cor(my_df$Sleep.Hours, my_df$Performance.Index)
cat("Correlation coefficient (r):", r, "\n")

# Calculate R-squared (r^2)
r_squared <- r^2
cat("Coefficient of determination (R²):", r_squared, "\n")

```
#Interpretation:
Very Weak Positive Correlation: The correlation is near zero, indicating almost no linear relationship between sleep hours and performance index.
R² = 0.0023: Only 0.23% of the variation in performance is explained by sleep duration.


#Do students who engage in extracurricular activities perform better or worse than those who don't?

```{r}
# Convert 'Extracurricular Activities' to binary (Yes = 1, No = 0)
my_df$Extracurricular.Activities <- ifelse(my_df$Extracurricular.Activities == "Yes", 1, 0)

# Calculate the average Performance Index for each group
mean_performance <- tapply(my_df$Performance.Index, my_df$Extracurricular.Activities, mean)
print("Mean Performance Index by Extracurricular Activities:")
print(mean_performance)

# Calculate correlation and R-squared
r <- cor(my_df$Extracurricular.Activities, my_df$Performance.Index)
r_squared <- r^2
```
#Extracurricular Activities on Performance:
Average Performance:

Students with extracurricular activities: 55.7

Students without extracurricular activities: 54.76

Correlation coefficient (r): 0.025

R²: 0.0006 (0.06% of the variation)

Interpretation:
Students who engage in extracurricular activities perform slightly better on average.
However, the correlation is extremely weak, suggesting that extracurricular involvement alone does not strongly predict performance.



#Is there an optimal balance between sleep, study time, and extracurricular participation?



# Load necessary package
install.packages("dplyr")
library(dplyr)

```{r}
# Convert 'Extracurricular Activities' to binary (Yes = 1, No = 0)
my_df$Extracurricular_Binary <- ifelse(my_df$Extracurricular.Activities == "Yes", 1, 0)
head(my_df)


# Build the multiple linear regression model
model <- lm(Performance.Index ~ Hours.Studied + Sleep.Hours + Extracurricular_Binary, data= my_df)

# Display the summary of the model
summary(model)



```
Multivariate Regression Summary:
We used Hours Studied, Sleep Hours, and Extracurricular Activities to predict Performance Index.

Key Findings:
Hours Studied (coef = 2.77, p < 0.001):

Each additional hour studied is associated with a 2.77 point increase in performance, on average.

This is the strongest predictor in the model.

Sleep Hours (coef = 0.55, p < 0.001):

Each extra hour of sleep contributes an average 0.55 point increase in performance.

Suggests sleep has a statistically significant but smaller effect than study time.


Model Performance:
R-squared = 0.143 → These three variables together explain 14.3% of the variance in performance.

The model is statistically significant (p < 0.001), but much of the variance is still unexplained, likely due to other factors (e.g., prior scores, study methods, motivation).


#	What combination of study hours, sleep, and extracurricular activities maximizes student performance?

```{r}
# Convert extracurricular participation to binary
my_df$Extracurricular_Binary <- ifelse(my_df$Extracurricular.Activities == "Yes", 1, 0) 

# Fit the linear regression model
model <- lm(Performance.Index ~ Hours.Studied + Sleep.Hours + Extracurricular.Activities, data = my_df)

# Extract coefficients
coeffs <- coef(model)
intercept <- coeffs["(Intercept)"]
coef_study <- coeffs["Hours.Studied"]
coef_sleep <- coeffs["Sleep.Hours"]
coef_extra <- coeffs["Extracurricular.Activities"]

# Initialize tracking variables
best_score <- -Inf #set to least possible value
best_combo <- c(NA, NA, NA)#creates a vector with the three placeholders

# Try all combinations
for (study in 0:10) 
    { #loop through hours 0-10
  for (sleep in 4:10) 
      { #loop through sleep 4-10
    for (extra in c(0, 1)) 
         { #loop though extra curricula, 0 or 1
      score <- intercept + coef_study * study + coef_sleep * sleep + coef_extra * extra
      if (!is.na(score) && score  > best_score) 
          {
        best_score <- score
        best_combo <- c(study, sleep, extra) #Update vector
           }
         }
       }
     }

# Display the best combination and predicted performance
cat("Best Combination:\n")
cat("Hours Studied:", best_combo[1], "\n")
cat("Sleep Hours:", best_combo[2], "\n")
cat("Extracurricular Participation:", ifelse(best_combo[3] == 1, "Yes", "No"), "\n")
cat("Predicted Performance Index:", round(best_score, 2), "\n")


```
According to the regression model, the highest predicted performance occurs when a student:

Studies 10 hours per day,

Sleeps 10 hours per night,

Participates in extracurricular activities.

While this combination is mathematically optimal, it may not be practically realistic due to time constraints. In reality, a balanced combination—such as 6–8 hours of study, 7–9 hours of sleep, and light extracurricular engagement—might be more sustainable.


```{r}
#Install packages
install.packages("ggcorrplot")
install.packages("reshape2")
# Load necessary libraries
library(ggplot2)
library(reshape2)
library(ggcorrplot)


# Define threshold for poor previous scores (bottom 25%)
low_score_threshold <- quantile(my_df$Previous.Scores, 0.25)

# Filter students with poor previous scores
low_score_students <- subset(my_df, Previous.Scores < low_score_threshold)

# Select only numeric columns
numeric_data <- low_score_students[, c("Hours.Studied", "Previous.Scores", 
                                       "Sleep.Hours", "Sample.Question.Papers.Practiced", 
                                       "Performance.Index")]

# Compute correlation matrix
cor_matrix <- cor(numeric_data, use = "complete.obs")

# Plot correlation heatmap
ggcorrplot(cor_matrix, lab = TRUE, colors = c("red", "white", "blue"),
           title = "Correlation Matrix for Students with Poor Previous Scores")

```
For students with poor previous scores (bottom 25%):

Hours Studied has a strong positive correlation with Performance Index (0.84).

Previous Scores still have some influence (0.46), suggesting baseline ability matters.

Sleep Hours show a weak positive correlation (0.11).

Sample Question Papers Practiced have minimal impact (0.06).


4. Segmentation and Grouping:
#Can students be grouped based on similar study habits and performance outcomes?

```{r}
# Load necessary libraries
library(dplyr)

# Read the dataset
df <- read.csv("Student_Performance.csv")

# Create performance group categories based on quantiles
low_threshold <- quantile(df$Performance.Index, 0.25)
high_threshold <- quantile(df$Performance.Index, 0.75)

df <- df %>%
  mutate(Performance.Group = cut(
    Performance.Index,
    breaks = c(-Inf, low_threshold, high_threshold, Inf),
    labels = c("Low", "Medium", "High")
  ))

# Calculate group-wise averages
group_summary <- df %>%
  group_by(Performance.Group) %>%
  summarise(
    Hours.Studied = round(mean(Hours.Studied), 2),
    Previous.Scores = round(mean(Previous.Scores), 2),
    Sleep.Hours = round(mean(Sleep.Hours), 2),
    Sample.Question.Papers.Practiced = round(mean(Sample.Question.Papers.Practiced), 2)
  )

# Display the summary
print(group_summary)
```
High-performing students studied significantly more hours (6.27 on average) and had much higher previous scores (90.5), along with slightly more sleep and practice.

Low-performing students studied the least (3.78 hours) and had much lower previous scores (49.1).

There’s a clear trend: more study time, slightly better sleep, and consistent practice all correlate with better performance.

5.Predictive Modeling:
#Can student performance be predicted accurately based on these factors?

```{r}
install.packages("readr")
install.packages("caret")
install.packages("dplyr")
install.packages("Metrics")
install.packages("ggcorrplot")

# Load necessary libraries
library(readr)
library(caret)
library(dplyr)
library(Metrics)
library(ggcorrplot, repos = "https://cloud.r-project.org/")


# Encode categorical variable
my_df$`Extracurricular_Activities` <- ifelse(my_df$`Extracurricular.Activities` == "Yes", 1, 0)

# Define features and target
features <- my_df %>% select(-`Performance.Index`)
target <- my_df$`Performance.Index`

# Split the data (80% train, 20% test)
set.seed(42)
train_index <- createDataPartition(target, p = 0.8, list = FALSE)
train_data <- features[train_index, ]
test_data <- features[-train_index, ]
train_target <- target[train_index]
test_target <- target[-train_index]

# Fit linear regression model
model <- lm(train_target ~ ., data = train_data)

# Predict on test data
predictions <- predict(model, newdata = test_data)

# Evaluate the model
r2 <- R2(predictions, test_target)
rmse_val <- rmse(test_target, predictions)

# Output results
cat("R²:", r2, "\n")
cat("RMSE:", rmse_val, "\n")

```
The linear regression model performs very well on this dataset:

R² Score: 0.989 — This means the model explains about 98.9% of the variance in the performance index.

RMSE (Root Mean Squared Error): ~2.02 — On average, the model’s predictions are about 2 points off from the actual performance index.

#What is the most influential factor in predicting student success?

```{r}
# Display coefficients sorted by influence
coef_summary <- summary(model)$coefficients
coefficients <- coef(model)[-1]  # Remove the intercept
sorted_coefficients <- sort(coefficients, decreasing = TRUE)

# Print sorted coefficients
print(sorted_coefficients)

# Optional: create a bar plot for visualization
barplot(sorted_coefficients,
        main = "Influence of Factors on Student Performance",
        col = "skyblue",
        las = 2,
        cex.names = 0.8,
        ylab = "Coefficient (Impact on Performance Index)")

```
#Based on the linear regression model, the most influential factors in predicting student success (measured by Performance Index) are:

Hours Studied – 2.85

Previous Scores – 1.02

Extracurricular Activities – 0.61

Sleep Hours – 0.48

Sample Question Papers Practiced – 0.19

#Interpretation:
Hours Studied has the strongest positive influence on performance.

Previous Scores are also a strong predictor, indicating consistent academic performance.

Extracurricular Activities and Sleep Hours contribute positively but to a lesser extent.

Practicing sample papers has the smallest effect among the factors listed.


